### 2022-05-31 doc/VerticalArrangements.tex
Finished updated simulation numbers to reflect bugfixes in code.

Our simulations focus on mergers that are more likely to have competitive effects and to therefore come under agency scruitiny. For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger. Similarly, when simulating a vertical merger, we assign the products sold by the largest wholesaler and the largest retailer to a single firm post-merger.

@@

The median average wholesale pre-merger price is \$4.50, and the median average retail pre-merger price is \$13.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.  Pre-merger HHIs range from 2,923 at the $25^{th}$ percentile to 4,966 at the $75^{th}$, with a median of 3,742. HHIs for vertical mergers increase by 1,298 points at the median, resulting in a median post-merger HHI equal to 5,995.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  HHIs for downstream mergers increase by 2,015 points at the median, resulting in a median post-merger HHI equal to 4,914.  HHIs for upstream mergers increase by 2,063 points at the median, resulting in a median post-merger HHI equal to 5,238. HHIs for integrated mergers increase by 2,770 points at the median, resulting in a median post-merger HHI equal to 7,280.

@@

We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.  Across all four types of mergers, the majority of cases show harm, though the distributions and magnitudes differ.  In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.   Median consumer harm from integrated mergers is almost 14\% of pre-merger total expenditures, 1.25 times the magnitude of that from downstream mergers, 2.3 times the magnitude of that from upstream mergers, and more than 4 times the magnitude of vertical mergers.

Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas downstream mergers, vertical mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 67\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.

Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 77\% of all upstream mergers, 23\% of vertical mergers, 20\% of downstream mergers, and about 15\% of integrated mergers. However, only wholesaler surplus from upstream mergers first-order stochastically dominates the wholesaler surplus of the other merger types.

As for total welfare, approximately 18\% of vertical mergers and 15\% of integrated mergers are beneficial, whereas only 3\% of upstream mergers and 2\% of downstream mergers are beneficial. Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream mergers, which dominates totals harm from vertical mergers. Median consumer harm from integrated mergers is about 9.6\% of pre-merger total expenditures, 1.5 times the magnitude of that from downstream mergers, 2.5 times the magnitude of that from upstream mergers, and more than 4.3 times the magnitude of vertical mergers.

@@

Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW} depicts paired box and whisker plots summarizing consumer harm (blue/darker bars on the left in each pair) and total harm (orange/lighter bars on the right in each pair) as the number of incumbent integrated firms increases. The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties. For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.

We begin with vertical mergers, as they are our simplest case, because there is never any preexisting integration at either merging firm.  Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about 36\% of vertical mergers benefit consumers and about 23\% of vertical mergers are net beneficial in total surplus.  The addition of a single rival integrated firm (1 integrated firm in the figure) narrows the distribution of outcomes slightly, but leaves median harm basically the same. Adding more integrated rivals tends to shift some outcomes from the lower tail to the upper tail, increasing the proportion of mergers with benefits, especially for consumers.  However, the inter-quartile range and median are largely unchanged.  Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers, though not dramatically.  Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.  It appears that these rivals can therefore somewhat limit the extent of negative impacts that would otherwise occur through RRC.

Moving to the left-most panel of Figure \ref{fig:CVvertincumbBW}, downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues. \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.  Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 18\% of mergers, and total welfare increases in 5.7\%, but median consumer harm grows to 14\% of pre-merger revenues, and median total harm grows to 7.3\%. Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.  On net, we find that the harms dominate in most instances.  Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat decreases the range of outcomes and moves the distribution towards less harm.  With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 27\% of markets and shrinks median consumer harm to 6.4\% of pre-merger revenues. The inter-quartile range of total harm narrows, and median total harm falls to about 3.8\% of pre-merger revenues. As we saw with vertical mergers, it appears that the presence of third party integrated rivals can limit harm, though here the effect is a bit stronger.

Next we examine the welfare impacts of upstream mergers.  Like with downstream mergers, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues. This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.  Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, but median harm is largely unchanged.  Consumer welfare increases in almost 16\% of mergers and total welfare increases in about 4.3\% of mergers, but median consumer harm and total harm are roughly steady at approximately 6.3\% and 4.8\% of pre-merger revenues, respectively.  The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.  Just as we saw with downstream mergers, here we find that harm dominates in most simulations.  Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- and slightly shrinks median harm.  Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.

We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW}.  Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 7\% of simulated markets and are net beneficial in 15\% of simulated markets. Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.  While adding one additional rival incumbent integrated firm (3 integrated firms in the figure) has little effect on the distribution of either consumer or total welfare, adding a second integrated rival (4 integrated firms in the figure) eliminates the most harmful integrated mergers by shortening the lower whisker. Adding further integrated incumbents has little incremental effect on the distribution of harm.

@@

For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.  Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargnobothBW} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW}.  Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power. Likewise, median total harm moves from about 2\% of pre-merger expenditures to 7.6\%.  In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.  Thus, there is little to counteract harms from decreased downstream competition.

Once we allow one of the merging firms to be integrated (denoted by 1 integrated firm in the figure), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.  The box and whisker plots rotate clockwise around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of harm.  Median consumer harm goes from approximately -33\% of pre-merger revenues when wholesalers have the advantage to 25\% of pre-merger revenues when retailers have the advantage.  Median total harm goes from about 1\% to 12.2\% of pre-merger revenues.  The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.  Compared to vertical mergers, where impacts to consumers were roughly neutral at equal bargaining power, here neutrality happens at 7/3, when wholesalers have a bit more than twice the power of retailers.  It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.  This is intuitive, as here the downstream merger also causes an additional lessening in horizontal competition.  Once we add rival integrated firms (see 4 integrated firms in the figure) the box and whisker plots rotate counter-clockwise, weakening but not erasing the relationship between bargaining power and harm. The range of outcomes also shrinks.  Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -33\% of pre-merger revenues to 14\% of pre-merger revenues, while median total harm goes from approximately 3.99\% to 5.5\% of pre-merger revenues.

We turn next to upstream mergers in Figure \ref{fig:CVbargnobothBW}.  Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.  Median consumer harm moves from about 29\% of pre-merger expenditures when wholesalers are more powerful to 2.2\% of pre-merger expenditures when retailers are more powerful. Likewise, median total harm goes from approximately 12\% to 0.7\% of pre-merger expenditures. By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figure), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.  This is intuitive, as now all three types of mergers create some form of vertical integration.  Median consumer harm goes from roughly -24\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 8.4\% of pre-merger expenditures when retailers have relatively more bargaining power. Median total harm increases from about -8.4\% to 3.5\% of pre-merger expenditures. Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.

Figure \ref{fig:CVbargbothBW} displays results for integrated mergers.  In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers. Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm. Median consumer harm moves from about 3.75\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power. Likewise, median total harm moves from about -14.5\% when wholesalers have relatively more bargaining power to 16\% of pre-merger expenditures when retailers have relatively more bargaining power.  A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 4 or 6 integrated firms in the figure) has little impact on the relationship between bargaining power and harm. However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 or more. The crossing at zero harm is shifted to the left.

@@

Our simulations confirm this hypothesis.  See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.  Overall, harm is greater for mergers with linear costs. The difference is largest for vertical mergers.  When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 30\% of simulations and total surplus in about 18\% of simulations, but only increase consumer surplus in about 1\% of simulations and total surplus in about 0.9\% of simulations when all firms have linear costs.

@@

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{Competitive Impact Statement: U.S. and Plantiff States v. Waste Management, Inc. and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impcat Statement: U.S. and State of Alabama v. Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{Waste Business Journal,\tiny{ \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anticompetitive effects were alleged for four SCCW collection markets and two MSW disposal markets. In addition, vertical anticompetitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements. The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.

@@

The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commerical waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers. The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors. In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere. Thus, pre-merger both parties were large, vertically integrated competitiors in the Chattanooga Area.

>>>>

Our simulations focus on mergers that are more likely to have competitive effects and to therefore come under agency scrutiny. For horizontal merger simulations, we assign the products sold by the two largest firms in the market to a single firm post-merger. Similarly, when simulating a vertical merger, we assign the products sold by the largest wholesaler and the largest retailer to a single firm post-merger.

@@

The median average wholesale pre-merger price is \$5.20, and the median average retail pre-merger price is \$14.  Recall that the market size is set to 1, meaning that average price is equal to total pre-merger expenditures.  Pre-merger HHIs range from 2,923 at the $25^{th}$ percentile to 4,966 at the $75^{th}$, with a median of 3,742. HHIs for vertical mergers increase by 1,328 points at the median, resulting in a median post-merger HHI equal to 6,046.\footnote{We compute the post-merger HHI for vertical mergers by treating the merged firms' market share as the sum of all the shares of downstream products that either use the upstream partner's input or are sold by the downstream partner.  }  HHIs for downstream mergers increase by 2,044 points at the median, resulting in a median post-merger HHI equal to 4,916.  HHIs for upstream mergers increase by 2,023 points at the median, resulting in a median post-merger HHI equal to 4,916. HHIs for integrated mergers increase by 2,822 points at the median, resulting in a median post-merger HHI equal to 7,135.

@@

We focus first on the results for consumers in the left-most panel of Figure \ref{fig:surplussum}.  Less than half of all simulated vertical mergers are harmful, while the majority of upstream, downstream, and integrated mergers show harm.  In particular, there is a partial rank-ordering of consumer harm across types of mergers: consumer harm from integrated mergers first-order stochastically dominates consumer harm from all other merger types, while downstream and upstream mergers each first-order stochastically dominate consumer harm from vertical mergers, but do not stochastically dominate one other.   Median consumer harm from integrated mergers is 14\% of pre-merger total expenditures, 1.45 times the magnitude of that from downstream mergers, and 3.1 times the magnitude of that from upstream mergers.

Moving on to retailers in the second panel of Figure \ref{fig:surplussum}, we find that whereas vertical mergers, downstream mergers, and integrated mergers always benefit retailers, upstream mergers harm retailers in about 60\% of all simulations. Moreover, there is a partial rank-ordering across mergers that is distinct from the consumer rank-ordering: the retailer surplus distribution from integrated mergers first-order stochastically dominates that from downstream mergers, which dominates that from upstream mergers, but not the retailer surplus distribution from vertical mergers.

Turning to wholesaler surplus in the third panel of Figure \ref{fig:surplussum}, the effects seen there are largely reversed from those for retailers: wholesaler surplus increases in about 67\% of all upstream mergers, 11\% of vertical mergers, 18\% of downstream mergers, and about 17\% of integrated mergers. The rank-ordering across mergers is also the reverse of the retailer surplus distribution.

As for total welfare, approximately 29\% of vertical mergers and 16\% of integrated mergers are beneficial, whereas only 7\% of upstream mergers and 3\% of downstream mergers are beneficial. Moreover, there is a partial rank ordering of mergers, with total harm from downstream mergers first-order stochastically dominating total harm from upstream  mergers, which dominates total harm from vertical mergers. Median consumer harm from integrated mergers is about 9.5\% of pre-merger total expenditures, 1.6 times the magnitude of that from downstream mergers, 2.6 times the magnitude of that from upstream mergers, and more than 4.6 times the magnitude of vertical mergers.

@@

Here, we examine merger outcomes both with and without preexisting vertical integration among the merging firms and as the number of rival incumbent integrated firms varies.  Figure \ref{fig:CVvertincumbBW} depicts box and whisker plots summarizing consumer harm (top row) and total harm (bottom row) as the number of incumbent integrated firms increases. The plots when the number of incumbent integrated firms equals 0 correspond to the results depicted in Figure 1 of \citeN{ST2021} and are included as reference.\footnote{An important difference between the simulations depicted in Figure \ref{fig:CVvertincumbBW} and those in Figure 1 of \citeN{ST2021} is that here we do not include downstream markets where prices are set according to a second score auction.}  For vertical mergers, the plots when the number of  incumbent integrated firms equals 1 summarize market outcomes where an unintegrated wholesaler merges with an unintegrated retailer, a single third party is integrated, and any remaining rivals are unintegrated. By contrast, for downstream and upstream  mergers, the plots when the number of incumbent integrated firms equals 1 depict the outcome of a merger between an integrated firm and an unintegrated retailer or wholesaler, respectively, with no integrated third parties. For integrated mergers, the plots when the number of incumbent integrated firms equals 2 (the lowest number allowed in this instance) depict a merger between two integrated firms, again with no integrated rivals.

We begin with vertical mergers, as they are our simplest case, because there is never any pre-existing integration at either merging firm.  Absent incumbent integration by rivals (denoted by 0 integrated firms in Figure \ref{fig:CVvertincumbBW}), about half of vertical mergers benefit consumers and about 31\% of vertical mergers increase total surplus.  The addition of a single rival integrated firm (1 integrated firm in the figure) increases median consumer harm from 0.2\% to to 2\% while widening the distribution of outcomes slightly. By contrast, additional integrating rivals has little effect on median total harm. Adding more integrated rivals tends to increase the proportion of mergers that benefit consumers, while also increasing the proportion of mergers that reduce total surplus. Thus, the presence of rival integrated firms can limit harm, particularly consumer harm, from vertical mergers.  Integrated rivals have sources of bargaining leverage that mirror the additional leverage a vertical merger brings to the combining firms.  It appears that these rivals can therefore somewhat limit the extent of negative impacts that would otherwise occur through RRC.

Moving to the second panel from the left in Figure \ref{fig:CVvertincumbBW}, upstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the range we study, with median consumer harm at about 6\% of pre-merger revenues and median total harm at about 3\% of pre-merger revenues. This result is to be expected, as upstream mergers without additional integration do not have the potential countervailing effects for welfare that have been discussed in the context of downstream or vertical mergers.  Once one of the merging firms is allowed to be integrated (1 integrated firm in the figure) beneficial mergers appear, and median harm changes modestly.  Consumer welfare increases in almost 30\% of mergers and total welfare increases in about 10\% of mergers, but median consumer harm and total harm are roughly steady at approximately 4\% pre-merger revenues.  The additional vertical integration creates some opportunities for EDM to enhance welfare, but this must be balanced against the potential for RRC.  Just as we saw with downstream mergers, here we find that harm dominates in most simulations.  Adding in integration by rival firms (2-4 integrated firms in the figure) narrows the inter-quartile range and trims the lower whisker-- meaning that relatively more harmful mergers are eliminated than beneficial ones-- while only slightly changing median harm.  Therefore, the presence of third-party incumbent integrated firms, as in the vertical and downstream cases, can limit the likelihood of observing extremely harmful mergers.

Next we examine the welfare impacts of downstream mergers. As with upstream mergers,  downstream mergers absent incumbent vertical integration (0 integrated firms in the figure) are never beneficial in the $5^{th}$ to $95^{th}$ percentile range, with median consumer harm from downstream mergers equal to about 9\% of pre-merger revenues and total harm equal to about 5\% of pre-merger revenues. \citeN{ST2021} found the same result for mergers in downstream logit Bertrand markets.  Allowing one of the merging firms to be vertically integrated when all other market participants are unintegrated (1 integrated firm in the figure) increases the range of outcomes while also leading to more median harm: consumer welfare increases in about 26\% of mergers, and total welfare increases in about 6\%, but median consumer harm grows to about 11\% of pre-merger revenues, and median total harm grows to about 7\%. Once one of the merging firms is already integrated, the merger now has potential EDM and RRC effects, which raises the possibility of both benefits and harms to welfare.  On net, we find that the harms dominate in most instances.  Adding integrated third party rivals (2-4 integrated firms in the figure) somewhat decreases the range of outcomes and moves the distribution towards less harm.  With 3 integrated third parties (4 integrated firms in the figure), the merger benefits consumers in about 34\% of markets and shrinks median consumer harm to about 5\% of pre-merger revenues. The inter-quartile range of total harm narrows, and median total harm falls to about 4\% of pre-merger revenues. As we saw with vertical mergers, it appears that the presence of third party integrated rivals can limit harm, though here the effect is a bit stronger.

We finish by examining integrated mergers in the right-most panel of Figure \ref{fig:CVvertincumbBW}.  Absent the presence of rival incumbent integrated firms (2 integrated firms in the figure), mergers between two integrated firms benefits consumers in about 8\% of simulated markets and are net beneficial in 15\% of simulated markets. Thus, we find that mergers between two firms that are already integrated are harmful in the vast majority of cases.  While adding one additional rival incumbent integrated firm (3 integrated firms in the figure) has little effect on the distribution of either consumer or total welfare, adding a second integrated rival (4 integrated firms in the figure) eliminates the most harmful integrated mergers by shortening the lower whisker. Adding further integrated incumbents has modest incremental effects on the distribution of harm.

@@

We turn next to upstream mergers in Figure \ref{fig:CVbargnobothBW}.  Without incumbent integrated firms in the market (0 integrated firms), the relationship between harm and bargaining power is reversed relative to downstream and vertical mergers: harm is greater when wholesalers have relatively more bargaining power.  Median consumer harm moves from about 28\% of pre-merger expenditures when wholesalers are more powerful to 2\% of pre-merger expenditures when retailers are more powerful. Likewise, median total harm goes from approximately 12\% to less than 1\% of pre-merger expenditures. By contrast, for mergers between an integrated and an unintegrated upstream supplier (denoted by 1 integrated firm in the figure), the relationship between harm and bargaining power is more in line with that for downstream and vertical mergers.  This is intuitive, as now all three types of mergers create some form of vertical integration.  Median consumer harm goes from roughly -23\% of pre-merger expenditures when wholesalers have relatively more bargaining power to almost 8\% of pre-merger expenditures when retailers have relatively more bargaining power. Median total harm increases from about -20\% to more than 3\% of pre-merger expenditures. Adding incumbent integrated rivals (see 4 integrated firms in the figure) preserves the relationship between bargaining power and harm, though the spread of outcomes narrows.

For downstream mergers, we continue to find that higher retailer relative bargaining power leads to more negative impacts on welfare.  Absent any incumbent integration (0 integrated firms), all the outcomes shown in Figure \ref{fig:CVbargnobothBW} feature a decrease in welfare, which mirrors our finding in Figure \ref{fig:CVvertincumbBW}.  Median consumer harm goes from about 1\% of pre-merger expenditures when wholesalers have relatively more bargaining power to 18\% of pre-merger expenditures when retailers have relatively more bargaining power. Likewise, median total harm moves from about 2\% of pre-merger expenditures to about 8\%.  In situations where retailers already have more bargaining power, retailers are likely to have extracted significant surplus from wholesalers prior to the merger, which limits any potential benefits from increased bargaining leverage.  Thus, there is little to counteract harms from decreased downstream competition.

Once we allow one of the merging firms to be integrated (denoted by 1 integrated firm in the figure), some beneficial mergers appear, particularly when wholesalers have relatively more bargaining power.  The box and whisker plots rotate clockwise around a pivot at equal bargaining power, strengthening the relationship between bargaining power and extent of harm.  Median consumer harm goes from approximately -33\% of pre-merger revenues when wholesalers have the advantage to 24\% of pre-merger revenues when retailers have the advantage.  Median total harm goes from about 2\% to almost 12\% of pre-merger revenues.  The existence of pre-merger integration at one of the merging firms creates an opportunity for EDM, the benefits from which are likely to be largest when wholesalers have high bargaining power and therefore charge high pre-merger input prices.  Compared to vertical mergers, where impacts to consumers were roughly neutral at 2/3 bargaining power, here neutrality happens between  3/2 and 7/3, when wholesalers have a bit more than 1.5 times the power of retailers.  It appears that wholesalers must be relatively more powerful (and thus the likely gains from EDM larger) in order to generate net consumer benefits, compared to in a vertical merger.  This is intuitive, as here the downstream merger also causes an additional lessening in horizontal competition.  Once we add rival integrated firms (see 4 integrated firms in the figure) the box and whisker plots rotate counter-clockwise, weakening but not erasing the relationship between bargaining power and harm. The range of outcomes also shrinks.  Starting from when wholesalers have relatively more bargaining power and moving right, median consumer harm goes from approximately -35\% of pre-merger revenues to 13\% of pre-merger revenues, while median total harm goes from almost 3\% to 5\% of pre-merger revenues.

Figure \ref{fig:CVbargbothBW} displays results for integrated mergers.  In terms of bargaining power and resulting harm, integrated mergers are perhaps most similar to vertical mergers. Like vertical mergers, there is a strong monotonic relationship between bargaining power and harm. Median consumer harm moves from less than -2\% of pre-merger expenditures when wholesalers have relatively more bargaining power to about 26\% of pre-merger expenditures when retailers have relatively more bargaining power. Likewise, median total harm moves from about -20\% when wholesalers have relatively more bargaining power to 15\% of pre-merger expenditures when retailers have relatively more bargaining power.  A second similarity between vertical and integrated mergers is that the presence of rival integrated incumbents (see 4 or 6 integrated firms in the figure) has little impact on the relationship between bargaining power and harm. However, a key difference between vertical and integrated mergers is that while vertical mergers are often beneficial when wholesalers have relatively more bargaining power, integrated mergers are often harmful unless wholesaler relative bargaining power is roughly 4 or more. The crossing at zero harm is shifted to the left.

@@

Our simulations confirm this hypothesis.  See the top two panels in Figure \ref{fig:surplussumcost}, which compare a scenario where all firms have constant marginal costs to one where they all have linear marginal costs.  Overall, harm is greater for mergers with linear costs. The difference is largest for vertical mergers.  When all firms have constant marginal costs, vertical mergers increase consumer surplus in about 55\% of simulations and total surplus in about 29\% of simulations, but only increase consumer surplus in about 36\% of simulations and total surplus in about 2\% of simulations when all firms have linear costs.

@@

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{Competitive Impact Statement: U.S. and Plaintiff States v. Waste Management, Inc. and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impact Statement: U.S. and State of Alabama v. Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{Waste Business Journal,\tiny{ \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anti-competitive effects were alleged for four SCCW collection markets and two MSW disposal markets. In addition, vertical anti-competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements. The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.

@@

The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commercial waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers. The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significant competitors. In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere. Thus, pre-merger both parties were large, vertically integrated competitors in the Chattanooga Area.

In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, and was the parties' sole competitor in the MSW disposal market. Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated. Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consummation since the merger was completed before the filing of the CIS for the Republic-Santek merger. The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.
